The dynamics of starvation and recovery: predicting extinction risk, Damuth's Law, and Cope's rule

### 2017-12-13 NatureComm_REVISION/YKR_NatureComm-R6.tex
editorial revisions

\documentclass[twocolumn,preprintnumbers,amsmath,amssymb,superscriptaddress]{revtex4}
%\usepackage[pdftex]{graphicx}

@@

%\usepackage{epstopdf}
%\usepackage{array}
%\usepackage{tabularx}
%\usepackage{multirow}
\usepackage{color}
%\usepackage{multibox}
%\usepackage{rotating}
%\usepackage{lineno}
%\usepackage[left]{lineno}
%\usepackage[comma,sort&compress]{natbib}
%\usepackage{authblk}
%\usepackage{multicol}

%\bibliographystyle{ieeetr}

\usepackage{bibunits}

%\linenumbers
%\setlength\linenumbersep{3pt}

@@

\author{Justin D. Yeakel} \affiliation{School of Natural Sciences, University
  of California, Merced, Merced, CA 95340, USA}

\author{Christopher P. Kempes} \affiliation{The Santa Fe Institute, 1399 Hyde
  Park Road, Santa Fe, NM 87501, USA}

\author{Sidney Redner} \affiliation{The Santa Fe Institute, 1399 Hyde Park
  Road, Santa Fe, NM 87501, USA}

\title{The dynamics of starvation and recovery}%: Eco-evolutionary feedbacks}

\begin{abstract} %250 words
The eco-evolutionary dynamics of species are fundamentally linked to the energetic constraints of its constituent individuals. Of particular importance is the interplay between reproduction and the dynamics of starvation and recovery. To elucidate this interplay, we introduce a nutritional state-structured model that incorporates two classes of consumer: nutritionally replete, reproducing consumers, and undernourished, non-reproducing consumers. We obtain strong constraints on starvation and recovery rates by deriving allometric scaling relationships and find that population dynamics are typically driven to a steady state. Moreover, these rates fall within a `refuge' in parameter space, where the probability of population extinction is minimized. We also show that our model provides a natural framework to predict maximum mammalian body size by determining the relative stability of an otherwise homogeneous population to a competing population with altered percent body fat. This framework provides a principled mechanism for a selective driver of Cope's rule.
\end{abstract}

@@

\begin{bibunit}[unsrt]

  The behavioral ecology of all organisms is influenced by their energetic
  states, which directly impacts how they invest reserves in uncertain
  environments.  Such behaviors are generally manifested as tradeoffs between
  investing in somatic maintenance and growth, or allocating energy towards
  reproduction~\citep{Martin:1987dl,Kirk:1997cc,Kempes:2012hy}.  The timing of
  these behaviors responds to selective pressure, as the choice of the
  investment impacts future
  fitness~\citep{Mangel:1988uaa,Mangel:2014kz,Yeakel:2013hi}.  The influence of
  resource limitation on an organism's ability to maintain its nutritional
  stores may lead to repeated delays or shifts in reproduction over the course
  of an organism's life.

@@

  Including individual energetic dynamics~\citep{Kooi2000} in a
  population-level framework~\citep{Kooi2000,Sousa:2010ez} is
  challenging~\citep{Diekmann:2010da}.  A common simplifying approach is the
  classic Lotka-Volterra (LV) model, which assumes that consumer population
  growth rate depends linearly on resource density~\citep{murdoch:2003}. Here,
  we introduce an alternative approach---the Nutritional State-structured Model
  (NSM)---that accounts for resource limitation via explicit starvation. In
  contrast to the LV model, the NSM incorporates two consumer states: hungry
  and full, with only the former susceptible to mortality and only the latter
  possessing sufficient energetic reserves to reproduce.  Additionally, we
  incorporate allometrically derived constraints on the time scales for
  reproduction~\citep{Kempes:2012hy}, starvation, and recovery.  Our model
  makes several basic predictions: (i) the dynamics are typically driven to a
  refuge far from cyclic behavior and extinction risk, (ii) the steady-state
  conditions of the NSM accurately predict the measured biomass densities for
  mammals described by Damuth's law~\citep{Damuth:1987kr,allen2002,enquist1998,Pedersen:2017he},
    %  energetic equivalence principle, 
  (iii) there is an allometrically constrained upper-bound for mammalian body size, and
  (iv) the NSM provides a selective mechanism for the evolution of larger body size, known as Cope's rule~\citep{Alroy:1998p1594,Clauset:2009fh,Smith:2010p3442,Saarinen:2014br}.

@@

  % In this work, we adopt an alternative approach to account for resource
  % limitation through the consequences of starvation.  Namely, in contrast to
  % the Lotka-Volterra model, we incorporate two states: hungry individuals and
  % individuals with sufficient energetic reserves to reproduce.}  Such a
  % constraint leads to reproductive time lags due to some members of the
  % population going hungry and then recovering.  Additionally, we incorporate
  % the idea that reproduction is strongly constrained allometrically
  % \citep{Kempes:2012hy}, and is not generally linearly related to resource
  % density.
  % As we shall show, these constraints influence the ensuing population dynamics in dramatic ways.\\

  \noindent \paragraph*{{\bf Nutritional state-structured model (NSM).}}
  We begin by defining the nutritional state-structured population model, where the consumer population is partitioned into two states: (a) an
  energetically replete (full) state $F$, where the consumer reproduces at a
  constant rate $\lambda$ and does not die from starvation, and (b) an
  energetically deficient (hungry) state $H$, where the consumer does not
  reproduce but dies by starvation at rate $\mu$. The dynamics of the
  underlying resource $R$ are governed by logistic growth with an intrinsic
  growth rate $\alpha$ and a carrying capacity $C$. The rate at which consumers
  transition between states and consume resources is dependent on their number,
  the abundance of resources, the efficiency of converting resources into
  metabolism, and how that metabolism is partitioned between maintenance and
  growth purposes.  We provide a physiologically and energetically mechanistic
  model for each of these dynamics and constants (see the Supplementary Information
  (SI)), and show that the system produces a simple non-dimensional form which
  we describe below.

  \begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{fig_FixedPoint2-eps-converted-to.pdf}
  \caption{\small{ The transcritical (TC; dashed line) and Hopf bifurcation (solid line) as a
    function of the starvation rate $\sigma$ and recovery rate $\rho$ for a 100g consumer.  These
    bifurcation conditions separate parameter space into unphysical (left of the TC), cyclic,
    and steady state dynamic regimes.  The colors show the steady state densities for the energetically replete consumers $F^*$.  
    % Steady state densities for the energetically deficient consumers $H^*$ (not shown)
    % scale with those for $F^*$.  
  }\label{fig:fp}}
  \end{figure}

  Consumers transition from the full state $F$ to the hungry state $H$ at a
  rate $\sigma$---the starvation rate---and also in proportion to the absence
  of resources $(1-R)$ (the maximum resource density has been non
  dimensionalized to 1; see SI).  Conversely, consumers recover from state $H$
  to state $F$ at rate $\xi \rho$ and in proportion to $R$, where $\xi$
  represents a ratio between maximal resource consumption and the carrying
  capacity of the resource. %and $\approx 2$.
  The resources that are eaten by hungry consumers (at rate $\rho R + \delta$)
  account for their somatic growth ($\rho R$) and maintenance ($\delta$).  Full
  consumers eat resources at a constant rate $\beta$ that accounts for maximal
  maintenance and somatic growth (see the SI for mechanistic derivations of
  these rates from resource energetics).
  %This inequality accounts for hungry consumers requiring more resources to replace lost body tissue.
  The NSM represents an ecologically motivated fundamental extension of the
  idealized starving random walk model of foraging, which focuses on resource
  depletion, to include reproduction and resource
  replenishment~\citep{Benichou:2014wu,Benichou:2016wl,Chupeau:2016jf}, and is
  a more general formulation than previous models that incorporate
  starvation~\citep{Persson:1998hz}.

@@

  This system of nondimensional equations follows from a set of first-principle
  relationships for resource consumption and growth (see the SI for a full derivation and the dimensional form).
  Notice that the total consumer density $F+H$ evolves according to $\dot{F}+\dot{H}=\lambda F-\mu H$.
  This resembles the equation of motion for the predator density in the LV model~\citep{murray2011mathematical}, except that the resource density does not appear in the growth term.
  % As discussed above, the attributes of reproduction and mortality have been explicitly apportioned to the full and hungry consumers, respectively, so that the growth in the total density is decoupled from the resource density.
  The rate of reproduction is independent of resource density because the full
  consumer partitions a constant amount of energy towards reproduction, whereas
  a hungry consumer partitions no energy towards reproduction.  Similarly, the
  consumer maintenance terms ($\delta H$ and $\beta F$) are also independent of
  resource density because they represent a minimal energetic requirement for
  consumers in the $H$ and $F$ state, respectively.

@@

  \noindent \paragraph*{ {\bf Steady states of the NSM.}} From the single
  internal fixed point (Eq.~\eqref{eq:ss}, see Methods), an obvious constraint
  on the NSM is that the reproduction rate $\lambda$ must be less than the
  starvation rate $\sigma$, so that the consumer and resource densities are
  positive.
  %In fact, when the resource density $R=0$, the rate equation for $F$ gives exponential growth of $F$ for $\lambda>\sigma$.
  The condition $\sigma = \lambda$ represents a transcritical (TC)
  bifurcation~\citep{Strogatz:2008wo} that demarcates a physical from an
  unphysical (negative steady-state densities) regime.  The biological
  implication of the constraint $\lambda<\sigma$ has a simple
  interpretation---the rate at which a macroscopic organism loses mass due to
  lack of resources is generally much faster than the rate of reproduction.  As
  we will discuss below, this inequality is also a natural consequence of
  allometric constraints~\citep{Kempes:2012hy} for organisms within empirically
  observed body size ranges. %(Fig.~\ref{fig:gvs})

@@

  \begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{Growth-trajectory-diagram-eps-converted-to.pdf}
  \caption{\small{ The growth trajectory over absolute time of an individual organism as a function of body mass.  
  Initial growth follows the black trajectory to an energetically replete reproductive adult mass of $m=\epsilon_\lambda M$ (see Methods). %which we assume is 95\% asymptotic mass $M$.  
  Starvation follows the red trajectory to $m = \epsilon_\sigma \epsilon_\lambda  M$. 
  Recovery follows the green curve to the replete adult mass, where this trajectory differs from the original growth because only fat is being regrown which requires a longer time to reach $\epsilon_\lambda M$. %different energetics and 
  Alternatively, death from starvation follows the blue trajectory to $m=\epsilon_\mu \epsilon_\lambda  M$.}\label{fig:growth}}
  \end{figure}

  % \noindent {\bf Role of allometry} \\
  %[Link Allometry stuff to our model - what does it provide to the story]
  %[Introduce biological and linked constraints]
  %[Allometries capture vast amounts of diversity via a single parameter --- body size]
  %[Allometries have captured intterest etc across multiple scales and biological classes of organisms]
  \noindent {\bf Results}
  \noindent \paragraph*{{\bf The allometry of extinction risk.}} While there
  are no {\it a priori} constraints on the parameters in the NSM, we expect
  that each species should be restricted to a distinct portion of the parameter
  space.  We use allometric scaling relations to constrain the covariation of
  rates in a principled and biologically meaningful manner (see Methods).
  Allometric scaling relations highlight common constraints and average trends
  across large ranges in body size and species diversity. Many of these
  relations can be derived from a small set of assumptions.  In the Methods we
  describe our framework to determine the covariation of timescales and rates
  across a range of body sizes for each of the key parameters of our model
  (cf.\ Ref.~\citep{Yodzis:1992hg}).
  % We are thereby able to define the regime of dynamics occupied by the entire class of mammals, along with the key differences between the largest and smallest mammals.

  Nearly all of the rates described in the NSM are determined by consumer
  metabolism, which can be used to describe a variety of organismal features
  \citep{Brown:2004wq}.  We derive, from first principles, the relationships
  for the rates of reproduction, starvation, recovery, and mortality as a
  function of an organism's body size and metabolic rate (see Methods).
  Because we aim to explore the starvation-recovery dynamics as a function of
  an organism's body mass $M$, we parameterize these rates in terms of the
  \emph{percent} gain and loss of the asymptotic (maximum) body mass,
  $\epsilon M$, where different values of $\epsilon$ define different states of
  the consumer (Fig.~\ref{fig:growth}; see Methods for derivations of
  allometrically constrained rate equations).  Although the rate equations
  \eqref{eq:system} are general and can in principle be used to explore the
  starvation recovery dynamics for most organisms, here we focus on allometric
  relationships for terrestrial-bound lower-trophic level endotherms (see the
  SI for values), specifically herbivorous mammals, which range from a minimum
  of $M\approx1$g (the Etruscan shrew \emph{Suncus etruscus}) to a maximum of
  $M\approx10^7$g (the early Oligocene Indricotheriinae and the Miocene
  Deinotheriinae).  Investigating other classes of organisms would simply
  involve altering the metabolic exponents and scalings associated with
  $\epsilon$. Moreover, we emphasize that our allometric equations (see
  Methods) describe mean relationships, and do not account for the (sometimes
  considerable) variance associated with individual species.  We note that
  including additional allometrically-scaled mortality terms to both $F$ and
  $H$ does not change the form of our model nor impact our quantitative findings
  (see SI for the derivation).

@@

  \begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{fig_ExtinctionAllometricComb4-eps-converted-to.pdf}
  \caption{\small{ Probability of extinction for a consumer with ({\bf a}) $M=10^2$g and ({\bf b}) $M=10^6$g as a function of the starvation rate $\sigma$ and recovery rate $\rho$, where the initial density is given as $(XF^*,XH^*,R^*)$, where $X$ is a random uniform variable in $[0,2]$. Note the change in scale in panel {\bf b}.  Extinction is defined as the population trajectory falling below $0.2\times$ the allometrically constrained steady state. The white points denote the allometrically constrained starvation and recovery rate.}\label{fig:ext}}
  \end{figure}

@@

  % \noindent \paragraph*{{\bf Extinction refuge.}} 
  We find that the allometrically constrained values of $\sigma$ and $\rho$,
  each representing different trajectories along the ontogenetic curve
  (Fig. \ref{fig:growth}), fall squarely within the extinction refuge across a
  range of $M$ (Fig. \ref{fig:ext}a,b, white points). These values are close
  enough to the Hopf bifurcation to avoid low steady-state densities, yet
  distant enough to avoid large-amplitude transient cycles.  Allometric values
  of $\sigma$ and $\rho$ fall within this relatively small window, which
  supports the possibility that a selective mechanism has constrained the
  physiological conditions driving starvation and recovery rates within
  populations.  Such a mechanism would select for organism physiology that
  generates appropriate $\sigma$ and $\rho$ values that minimize extinction
  risk.  This selection could occur via the tuning of body fat percentages,
  metabolic rates, and/or biomass maintenance efficiencies.  We also find that
  as body size increases, the size of the low extinction-risk parameter space
  shrinks (Fig.~\ref{fig:ext}b), suggesting that the population dynamics for
  larger organisms are more sensitive to variability in physiological rates.
  This finding is in accordance with, and may serve as contributing support for, observations of increased extinction risk among larger mammals \cite{Liow:2008jx}.\\
  % Moreover, larger body sizes decrease the steady state resource density, such that fluctuations for larger organisms will be more likely to drive resources to extinction.
  % OLD WORDING Such a mechanism would involve a feedback between the dynamics of the population and the fitness of individuals within the population, though to what extent the dynamics of the population influence rates of starvation and recovery would also involve potential tradeoffs in reproduction and somatic maintenance.
  % To summarize, our finding that allometrically-determined parameters fall within this low extinction probability region suggests that the NSM dynamics may both drive---and constrain---natural animal populations.\\

@@

  % \noindent {\bf Energetic barriers to body size}
  \noindent \paragraph*{{\bf Damuth's Law and body size limits.}} The NSM
  correctly predicts that smaller species have larger steady-state population
  densities (Fig.~\ref{fig:mass}).  Similar predictions have been made for
  carnivore populations using alternative consumer-resource models
  \citep{DeLong:2012kw}.  Moreover, we show that the NSM provides independent
  theoretical support for Damuth's Law
  \citep{Damuth:1987kr,allen2002,enquist1998,Pedersen:2017he}.  Damuth's
  law shows that species abundances, $N^{*}$, follow $N^*=0.01
  M^{-0.78}$ (g m$^{-2}$). Figure \ref{fig:mass} shows that both $F^{*}$ and $H^{*}$ scale
  as $M^{-\eta}$, with $\eta\approx 3/4$,  over a wide range of organismal sizes and that $F^{*}+H^{*}$
  closely matches the best fit to Damuth's data.  Remarkably, this result
  illustrates that the steady state values of the NSM combined with the derived
  timescales naturally give rise to Damuth's law. While the previous metabolic
  studies supporting Damuth's law provided arguments for the value of the
  exponent~\citep{allen2002}, these studies are only able to infer the
  normalization constant ($0.01$ g$^{1.78}$ m$^{-2}$ in the above equation) from the data (see SI for a discussion of the energy
  equivalence hypothesis related to these metabolic arguments). Our model
  predicts not only the exponent but also the normalization constant by
  explicitly including the resource dynamics and the parameters that determine
  growth and consumption. It should be noted that density relationships of
  individual clades follow a more shallow scaling relationship than predicted
  by Damuth's law~\cite{Pedersen:2017he}.  In the context of our model,
  this finding suggests that future work may be able to anticipate these shifts
  by accounting for differences in the physiological parameters associated with
  each clade.

  \begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{fig_FPAllometric2-eps-converted-to.pdf}
  \caption{\small{Consumer steady states $F^*$ (green) and $H^*$ (orange) as a function of
    body mass along with the data from Damuth \citep{Damuth:1987kr}. Inset: Resource steady state $R^*$ as a function of consumer body mass.}\label{fig:mass}}
  \end{figure}

>>>>

% \RequirePackage{lineno}
\documentclass[twocolumn,preprintnumbers,amsmath,amssymb,superscriptaddress]{revtex4}
%\usepackage[pdftex]{graphicx}

@@

%\usepackage{epstopdf}
%\usepackage{array}
%\usepackage{tabularx}
%\usepackage{multirow}
\usepackage{color}
%\usepackage{multibox}
%\usepackage{rotating}
\usepackage{lineno}
%\usepackage[left]{lineno}
\usepackage[comma,sort&compress]{natbib}
\bibpunct{}{}{,}{s}{}{;}
%\usepackage{authblk}
%\usepackage{multicol}

\bibliographystyle{naturemag}

% \usepackage{bibunits}

% \linenumbers
% \setlength\linenumbersep{3pt}

@@

\author{Justin D. Yeakel,${}^{1,2,*}$, Christopher P. Kempes,${}^{2,*}$,Sidney Redner,${}^{2,*}$\\
${}^{1}$School of Natural Sciences, University of California, Merced, Merced, CA 95340, USA\\
${}^{2}$The Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, NM 87501, USA\\
${}^*$Contributed equally\\
Correspondence and requests for materials should be addressed to J.D.Y. (email: jdyeakel@gmail.com), C.P.K. (email: ckempes@gmail.com), S.R. (email: sredner@santafe.edu)
} 
% \affiliation{School of Natural Sciences, University
%   of California, Merced, Merced, CA 95340, USA}\affiliation{The Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, NM 87501, USA}\affiliation{Contributed equally}\affiliation{Corresponding author}

% \author{} 
% \affiliation{The Santa Fe Institute, 1399 Hyde
%   Park Road, Santa Fe, NM 87501, USA}\affiliation{Contributed equally}\affiliation{Corresponding author}

% \author{} 
% \affiliation{The Santa Fe Institute, 1399 Hyde Park
%   Road, Santa Fe, NM 87501, USA}\affiliation{Contributed equally}\affiliation{Corresponding author}
\title{The dynamics of starvation and recovery: predicting extinction risk, Damuth's Law, and Cope's rule}%: Eco-evolutionary feedbacks}

\begin{abstract} %250 words
The eco-evolutionary dynamics of species are fundamentally linked to the energetic constraints of their constituent individuals. Of particular importance is the interplay between reproduction and the dynamics of starvation and recovery. To elucidate this interplay, we here introduce a nutritional state-structured model that incorporates two classes of consumer: nutritionally replete, reproducing consumers, and undernourished, non-reproducing consumers. We obtain strong constraints on starvation and recovery rates by deriving allometric scaling relationships and find that population dynamics are typically driven to a steady state. Moreover, these rates fall within a `refuge' in parameter space, where the probability of population extinction is minimized. We also show that our model provides a natural framework to predict maximum mammalian body size by determining the relative stability of an otherwise homogeneous population to a competing population with altered percent body fat. This framework provides a principled mechanism for a selective driver of Cope's rule.
\end{abstract}

@@

% \begin{bibunit}[unsrt]

\noindent {\bf Introduction}\\
The behavioral ecology of all organisms is influenced by their energetic
states, which directly impacts how they invest reserves in uncertain
environments.  Such behaviors are generally manifested as tradeoffs between
investing in somatic maintenance and growth, or allocating energy towards
reproduction~\citep{Martin:1987dl,Kirk:1997cc,Kempes:2012hy}.  The timing of
these behaviors responds to selective pressure, as the choice of the
investment impacts future
fitness~\citep{Mangel:1988uaa,Mangel:2014kz,Yeakel:2013hi}.  The influence of
resource limitation on an organism's ability to maintain its nutritional
stores may lead to repeated delays or shifts in reproduction over the course
of an organism's life.

@@

Including individual energetic dynamics~\citep{Kooi2000} in a
population-level framework~\citep{Kooi2000,Sousa:2010ez} is
challenging~\citep{Diekmann:2010da}.  A common simplifying approach is the
classic Lotka-Volterra (LV) model, which assumes that consumer population
growth rate depends linearly on resource density~\citep{murdoch:2003}. Here,
we introduce an alternative approach---the Nutritional State-structured Model
(NSM)---that accounts for resource limitation via explicit starvation. In
contrast to the LV model, the NSM incorporates two consumer states: hungry
and full, with only the former susceptible to mortality and only the latter
possessing sufficient energetic reserves to reproduce.  Additionally, we
incorporate allometrically derived constraints on the time scales for
reproduction~\citep{Kempes:2012hy}, starvation, and recovery.  Our model
makes several basic predictions: (i) the dynamics are typically driven to a
refuge far from cyclic behavior and extinction risk, (ii) the steady-state
conditions of the NSM accurately predict the measured biomass densities for
mammals described by Damuth's law~\citep{Damuth:1987kr,allen2002,enquist1998,Pedersen:2017he},
  %  energetic equivalence principle, 
(iii) there is an allometrically constrained upper-bound for mammalian body size, and
(iv) the NSM provides a selective mechanism for the evolution of larger body size, known as Cope's rule~\citep{Alroy:1998p1594,Clauset:2009fh,Smith:2010p3442,Saarinen:2014br}.\\

@@

% In this work, we adopt an alternative approach to account for resource
% limitation through the consequences of starvation.  Namely, in contrast to
% the Lotka-Volterra model, we incorporate two states: hungry individuals and
% individuals with sufficient energetic reserves to reproduce.}  Such a
% constraint leads to reproductive time lags due to some members of the
% population going hungry and then recovering.  Additionally, we incorporate
% the idea that reproduction is strongly constrained allometrically
% \citep{Kempes:2012hy}, and is not generally linearly related to resource
% density.
% As we shall show, these constraints influence the ensuing population dynamics in dramatic ways.\\
\noindent {\bf Results}\\
\noindent {\bf Nutritional state-structured model (NSM)}\\
We begin by defining the nutritional state-structured population model, where the consumer population is partitioned into two states: (a) an
energetically replete (full) state $F$, where the consumer reproduces at a
constant rate $\lambda$ and does not die from starvation, and (b) an
energetically deficient (hungry) state $H$, where the consumer does not
reproduce but dies by starvation at rate $\mu$. The dynamics of the
underlying resource $R$ are governed by logistic growth with an intrinsic
growth rate $\alpha$ and a carrying capacity $C$. The rate at which consumers
transition between states and consume resources is dependent on their number,
the abundance of resources, the efficiency of converting resources into
metabolism, and how that metabolism is partitioned between maintenance and
growth purposes.  We provide a physiologically and energetically mechanistic
model for each of these dynamics and constants (see Methods), and show that the system produces a simple non-dimensional form which
we describe below.

Consumers transition from the full state $F$ to the hungry state $H$ at a
rate $\sigma$---the starvation rate---and also in proportion to the absence
of resources $(1-R)$ (the maximum resource density has been non
dimensionalized to 1; see Methods).  Conversely, consumers recover from state $H$
to state $F$ at rate $\xi \rho$ and in proportion to $R$, where $\xi$
represents a ratio between maximal resource consumption and the carrying
capacity of the resource. %and $\approx 2$.
The resources that are eaten by hungry consumers (at rate $\rho R + \delta$)
account for their somatic growth ($\rho R$) and maintenance ($\delta$).  Full
consumers eat resources at a constant rate $\beta$ that accounts for maximal
maintenance and somatic growth (see Methods for mechanistic derivations of
these rates from resource energetics).
%This inequality accounts for hungry consumers requiring more resources to replace lost body tissue.
The NSM represents an ecologically motivated fundamental extension of the
idealized starving random walk model of foraging, which focuses on resource
depletion, to include reproduction and resource
replenishment~\citep{Benichou:2014wu,Benichou:2016wl,Chupeau:2016jf}, and is
a more general formulation than previous models that incorporate
starvation~\citep{Persson:1998hz}.

@@

This system of nondimensional equations follows from a set of first-principle
relationships for resource consumption and growth (see Methods for a full derivation and the dimensional form).
Notice that the total consumer density $F+H$ follows $\dot{F}+\dot{H}=\lambda F-\mu H$.
This resembles the equation of motion for the predator density in the LV model~\citep{murray2011mathematical}, except that the resource density does not appear in the growth term.
% As discussed above, the attributes of reproduction and mortality have been explicitly apportioned to the full and hungry consumers, respectively, so that the growth in the total density is decoupled from the resource density.
The rate of reproduction is independent of resource density because the full
consumer partitions a constant amount of energy towards reproduction, whereas
a hungry consumer partitions no energy towards reproduction.  Similarly, the
consumer maintenance terms ($\delta H$ and $\beta F$) are also independent of
resource density because they represent a minimal energetic requirement for
consumers in the $H$ and $F$ state, respectively.\\

@@

\noindent {\bf Steady states of the NSM}\\ 
From the single
internal fixed point (Eq.~\eqref{eq:ss}, see Methods), an obvious constraint
on the NSM is that the reproduction rate $\lambda$ must be less than the
starvation rate $\sigma$, so that the consumer and resource densities are
positive.
%In fact, when the resource density $R=0$, the rate equation for $F$ gives exponential growth of $F$ for $\lambda>\sigma$.
The condition $\sigma = \lambda$ represents a transcritical (TC)
bifurcation~\citep{Strogatz:2008wo} that demarcates a physical from an
unphysical (negative steady-state densities) regime.  The biological
implication of the constraint $\lambda<\sigma$ has a simple
interpretation---the rate at which a macroscopic organism loses mass due to
lack of resources is generally much faster than the rate of reproduction.  As
we will discuss below, this inequality is also a natural consequence of
allometric constraints~\citep{Kempes:2012hy} for organisms within empirically
observed body size ranges. %(Fig.~\ref{fig:gvs})

@@

% \noindent {\bf Role of allometry} \\
%[Link Allometry stuff to our model - what does it provide to the story]
%[Introduce biological and linked constraints]
%[Allometries capture vast amounts of diversity via a single parameter --- body size]
%[Allometries have captured intterest etc across multiple scales and biological classes of organisms]

\noindent {\bf The allometry of extinction risk}\\
While there
are no {\it a priori} constraints on the parameters in the NSM, we expect
that each species should be restricted to a distinct portion of the parameter
space.  We use allometric scaling relations to constrain the covariation of
rates in a principled and biologically meaningful manner (see Methods).
Allometric scaling relations highlight common constraints and average trends
across large ranges in body size and species diversity. Many of these
relations can be derived from a small set of assumptions.  In the Methods we
describe our framework to determine the covariation of timescales and rates
across a range of body sizes for each of the key parameters of our model
(cf.\ Ref.~\citep{Yodzis:1992hg}).
% We are thereby able to define the regime of dynamics occupied by the entire class of mammals, along with the key differences between the largest and smallest mammals.

Nearly all of the rates described in the NSM are determined by consumer
metabolism, which can be used to describe a variety of organismal features
\citep{Brown:2004wq}.  We derive, from first principles, the relationships
for the rates of reproduction, starvation, recovery, and mortality as a
function of an organism's body size and metabolic rate (see Methods).
Because we aim to explore the starvation-recovery dynamics as a function of
an organism's body mass $M$, we parameterize these rates in terms of the
percent gain and loss of the asymptotic (maximum) body mass,
$\epsilon M$, where different values of $\epsilon$ define different states of
the consumer (Fig.~\ref{fig:growth}; see Methods for derivations of
allometrically constrained rate equations).  Although the rate equations
\eqref{eq:system} are general and can in principle be used to explore the
starvation recovery dynamics for most organisms, here we focus on allometric
relationships for terrestrial-bound lower-trophic level endotherms (table \ref{param}), specifically herbivorous mammals, which range from a minimum
of $M\approx1$g (the Etruscan shrew \emph{Suncus etruscus}) to a maximum of
$M\approx10^7$g (the early Oligocene Indricotheriinae and the Miocene
Deinotheriinae).  Investigating other classes of organisms would simply
involve altering the metabolic exponents and scalings associated with
$\epsilon$. Moreover, we emphasize that our allometric equations (see
Methods) describe mean relationships, and do not account for the (sometimes
considerable) variance associated with individual species.  We note that
including additional allometrically-scaled mortality terms to both $F$ and
$H$ does not change the form of our model nor impact our quantitative findings
(see Supplementary Note 1: Sensitivity to additional death terms).

@@

@@

% \noindent \paragraph*{{\bf Extinction refuge.}} 
We find that the allometrically constrained values of $\sigma$ and $\rho$,
each representing different trajectories along the ontogenetic curve
(Fig. \ref{fig:growth}), fall squarely within the extinction refuge across a
range of $M$ (Fig. \ref{fig:ext}a,b, white points). These values are close
enough to the Hopf bifurcation to avoid low steady-state densities, yet
distant enough to avoid large-amplitude transient cycles.  Allometric values
of $\sigma$ and $\rho$ fall within this relatively small window, which
supports the possibility that a selective mechanism has constrained the
physiological conditions driving starvation and recovery rates within
populations.  Such a mechanism would select for organism physiology that
generates appropriate $\sigma$ and $\rho$ values that minimize extinction
risk.  This selection could occur via the tuning of body fat percentages,
metabolic rates, and/or biomass maintenance efficiencies.  We also find that
as body size increases, the size of the low extinction-risk parameter space
shrinks (Fig.~\ref{fig:ext}b), suggesting that the population dynamics for
larger organisms are more sensitive to variability in physiological rates.
This finding is in accordance with, and may serve as contributing support for, observations of increased extinction risk among larger mammals \citep{Liow:2008jx}.\\
% Moreover, larger body sizes decrease the steady state resource density, such that fluctuations for larger organisms will be more likely to drive resources to extinction.
% OLD WORDING Such a mechanism would involve a feedback between the dynamics of the population and the fitness of individuals within the population, though to what extent the dynamics of the population influence rates of starvation and recovery would also involve potential tradeoffs in reproduction and somatic maintenance.
% To summarize, our finding that allometrically-determined parameters fall within this low extinction probability region suggests that the NSM dynamics may both drive---and constrain---natural animal populations.\\

@@

% \noindent {\bf Energetic barriers to body size}
\noindent {\bf Damuth's Law and body size limits}\\ 
The NSM
correctly predicts that smaller species have larger steady-state population
densities (Fig.~\ref{fig:mass}).  Similar predictions have been made for
carnivore populations using alternative consumer-resource models
\citep{DeLong:2012kw}.  Moreover, we show that the NSM provides independent
theoretical support for Damuth's Law
\citep{Damuth:1987kr,allen2002,enquist1998,Pedersen:2017he}.  Damuth's
law shows that species abundances, $N^{*}$, follow $N^*=0.01
M^{-0.78}$ (g m$^{-2}$). Figure \ref{fig:mass} shows that both $F^{*}$ and $H^{*}$ scale
as $M^{-\eta}$, with $\eta\approx 3/4$,  over a wide range of organismal sizes and that $F^{*}+H^{*}$
closely matches the best fit to Damuth's data.  Remarkably, this result
illustrates that the steady state values of the NSM combined with the derived
timescales naturally give rise to Damuth's law. 
While the initial metabolic studies supporting Damuth's law provide arguments for the value of the exponent \citep{Damuth:1987kr,allen2002} (see Supplementary Note 2: NSM and the energy equivalence hypothesis), our model predicts not only the exponent but also the normalization constant by explicitly including the resource dynamics and the parameters that determine growth and consumption. 
These predictions are complementary to recent work that also predicts the exponent and normalization constant of density relationships from the detailed allometries of reproduction, capture area, conversion efficiency, and mortality within predator-prey dynamic models \citep{DeLong:2012kw,DeLong:2012fj}.
It should be noted that density relationships of
individual clades follow a more shallow scaling relationship than predicted
by Damuth's law~\citep{Pedersen:2017he}.  In the context of our model,
this finding suggests that future work may be able to anticipate these shifts
by accounting for differences in the physiological parameters associated with
each clade.

% \paragraph*{{\bf Maximum Body Size}}
%JUSTIN I REMOVED: Our model shows that energetic equivalence breaks down at large $M$ suggesting that this maximum is a hard limit where deviations outside of this range are energetically suboptimal.  
%, violating the energy equivalence. % set by deviations from constant efficiency obeyed by other populations.
With respect to predicted steady state densities, the population sizes of F and H become infinite at a finite body size, where the steady state resources also vanish (Fig.~\ref{fig:mass}). 
This asymptotic behavior is governed by body sizes at which $\epsilon_{\rm \mu}$ and
$\epsilon_{\rm \lambda}$ (see Fig.~\ref{fig:growth}) equal zero, causing the
timescales (Eqn. \ref{t1}) to become infinite and the rates $\mu$ and $\lambda$ to equal
zero.
% A theoretical upper bound on mammalian body size is given by
% $\epsilon_\sigma=0$, where mammals are entirely composed of metabolic
% reserves, and this occurs at $M=8.3\times 10^8$ (g), or $120$ times the mass
% of a male African elephant.
The $\mu=0$ asymptote occurs first when
$f_{\rm 0}M^{\gamma-1}+u_{\rm 0}M^{\zeta-1}=1$, and corresponds to
$(F^*,H^*,R^*)=(0,0,0)$.  This point predicts an upper bound on mammalian
body size at $M_{\rm \rm max}=6.54\times10^7$ (g).  Moreover, $M_{\rm \rm max}$,
which is entirely determined by the population-level consequences of
energetic constraints, is within an order of magnitude of the maximum body
size observed in the North American mammalian fossil
record~\citep{Alroy:1998p1594}, as well as the mass predicted from an
evolutionary model of body size evolution~\citep{Clauset:2009fh}.  We
emphasize that the asymptotic behavior and predicted upper bound depend only
on the scaling of body composition and are independent of the resource
parameters.  The prediction of an asymptotic limit on mammalian size
parallels work on microbial life where an upper and lower bound on bacterial
size, and an upper bound on single cell eukaryotic size, is predicted from
similar growth and energetic scaling
relationships~\citep{Kempes:2012hy,Kempes:2016}.  It has also been shown that
models that incorporate the allometry of hunting and resting combined with
foraging time predicts a maximum carnivore size between $7\times10^{5}$ and
$1.1\times10^{6}$ (g) \citep{Carbone:1999ju,Carbone:2007dz}.  Similarly, the
maximum body size within a particular lineage has been shown to scale with
the metabolic normalization constant
\citep{Okie:2013ju}. %and depend on a critical death found to be constant from data
This complementary approach is based on the balance between growth and
mortality, and suggests that future connections between the scaling of fat
and muscle mass should systematically be connected with $B_{\rm 0}$ when
comparing
lineages.
